Human capital mediates natural selection in contemporary humans

### 2022-01-02 why-natural-selection.Rmd
Run parental AFLB tercile regressions. Don't plot respondents regression controlling for AFLB.

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related 
polygenic scores. Nevertheless, differences in equation
\@ref(eq:regression) across subgroups constrain the set of possible explanations
for natural selection. As we show below, the economic theory predicts 
these differences, while alternative explanations do not.

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is 
`r cor_age_flb`.

@@

To investigate this further, we estimate equation \@ref(eq:regression)
*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
controls are added. The correlation between effect sizes controlling for AFLB,
and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
through two opposing channels: an effect on AFLB, and an opposite-signed effect
on number of children controlling for AFLB. Again, we do not claim that AFLB
is exogenous to an individual's polygenic scores. Indeed, our analyses show
that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
that AFLB mediates part of the relationship between polygenic scores and
fertility, and that once this is controlled for, the remaining part of the
relationship has the opposite sign.

```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles (women only)", "Effect sizes controlling for age at first live birth. Effect sizes without controls (for women only) shown for comparison."), fig.align = "center"}

@@

n_regs <- as.double(nrow(res_age_flb))

res_combined %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs(2)),
        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = Regression, n_regs = n_regs)

@@

We also examine selection effects among respondents' parents, using
information on respondents' number of siblings. Effect sizes of polygenic
scores are highly correlated across the two generations. Median-splitting
respondents by year of birth, we find little evidence of change in effect sizes
among either the parents' or the respondents' generation. We check whether 
selection effects differ by socio-economic status in the
parents' generation. We have no information about parents' income, so we use the
1971 Townsend deprivation score of respondents' birthplace as a proxy
[@townsend1987deprivation]. Results show the same pattern as for respondents:
effect sizes are larger and more often significant in the most deprived areas
(Appendix Figure \@ref(fig:plot-siblings-townsend)); effects change
sign when controlling for AFLB (Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, the siblings data 
lets us check for a "quantity-quality tradeoff" between
number of children (of respondents' parents) and number of grandchildren. We
do not find any: in fact, the correlation between *number of siblings* and 
*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
$p < 2 \times 10^{-16}$).

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of risk aversion, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

[^one-period-results]: Predictions 1-3 also hold in the one-period model with
CRRA utility. Our empirical results are actually stronger than prediction 5, in
that correlations with fertility are *reversed* at higher AFLB. This prediction
can be accommodated in the model if children have a money cost as well as a time
cost. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.

@@

5.  The relationship is weaker among those            Effects weaker among those starting 
    who start fertility in period 2                   fertility later 
    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)a).
    start fertility in period 1 
    ($N_{1}^{*} > 0$).
----------------------------------------------------------------------------------------

@@

@@

\clearpage
\FloatBarrier
## Selection controlling for age at first live birth: respondents' parents

@@

Among the parents' generation, we can control for age at first live
birth using the subsets of respondents who reported their mother's or
father's age, and who had no elder siblings. We run regressions on
*number of siblings* on these subsets, controlling for either parent's
age at their birth. Figure \@ref(fig:plot-age-birth-parents) shows the
results. Effect sizes are very similar, whether controlling for father's
or mother's age. As in the respondents' generation, effect sizes
are negatively correlated with the effect sizes from bivariate
regressions without the age at birth control (father's age at birth: 
$\rho$ `r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).

>>>>

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.
We emphasize that these categories are not necessarily exogenous to the polygenic
scores. For example, both in the data and in our theoretical model, age at first
live birth is a choice variable, which is endogenous to human-capital-related
polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
differences in equation \@ref(eq:regression) across subgroups constrain the set
of possible explanations for natural selection. As we show below, a simple
economic model predicts the differences we find.

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is 
`r cor_age_flb`.

@@

To investigate this further, we estimate equation \@ref(eq:regression) among
females, *controlling* for AFLB. In `r opp_signed` out of 
`r length(score_names)` cases, effects change sign when controls are added. The
correlation between effect sizes controlling for AFLB, and raw effect sizes, is
`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
channels: an effect on AFLB, and an opposite-signed effect on number of children
after AFLB is controlled for.

```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth terciles (women only).", fig.align = "center"}

@@

@@

We also examine selection effects among respondents' parents, using information
on respondents' number of siblings. Effect sizes of polygenic scores are highly
correlated across the two generations. Median-splitting respondents by year of
birth, we find little evidence of change in effect sizes among either the
parents' or the respondents' generation. We also check whether selection effects
vary by socio-economic status and AFLB in the parents' generation. We have no
information about parents' income, so we use the 1971 Townsend deprivation score
of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
the same pattern as for respondents. Effect sizes are larger and more often
significant in the most deprived areas (Appendix Figure
\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
and change sign when controlling for AFLB, among both mothers and fathers
(Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, we check for a
"quantity-quality tradeoff" between parents' number of children and number of
grandchildren. We do not find any: in fact, the correlation between *number of
siblings* and *number of children* is positive ($\rho$ = 
`r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).

@@

Instead of identifying human capital with wages, we now allow
individuals to choose a level of education $s \in [0,1]$, which has
a time cost in period 1. Education is
complementary to human capital $h > 0$, and increases period 2 wages, which take 
the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
predictions, along with corresponding empirical results for the correlation
between polygenic scores and fertility.[^one-period-results] The key insight of
the model is that for middling levels of risk aversion, the substitution effect
dominates at low income levels, but is balanced by the income effect as income
increases.

[^one-period-results]: Predictions 1-3 also hold in the one-period model with
constant relative risk aversion. Our empirical results are actually stronger
than prediction 5, in that correlations with fertility are *reversed* at higher
AFLB. This prediction can be accommodated in the model if children have a money
cost as well as a time cost (Appendix Figure \@ref(fig:N-plot-with-m)).

@@

5.  The relationship is weaker among those            Effects weaker among those starting 
    who start fertility in period 2                   fertility later 
    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)).
    start fertility in period 1 
    ($N_{1}^{*} > 0$).
----------------------------------------------------------------------------------------

@@

\clearpage
\FloatBarrier
## Parents: area deprivation

@@

\clearpage
\FloatBarrier
## Parents: age at first live birth

@@

Among the parents' generation, we can control for age at first live birth using
the subsets of respondents who reported their mother's or father's age, and who
had no elder siblings. We run regressions on *number of siblings* on these
subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
by terciles of age at first live birth, for mothers and fathers. As in the
respondents' generation, effect sizes are larger for younger parents. This
holds for both sexes.

```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}

drake::loadd(res_age_flb_mothers_cross)
drake::loadd(res_age_flb_fathers_cross)

famhist_bo1 <- famhist %>% 
                 filter(birth_order == 1) %>%
                 select(n_siblings, whr_combined, moth_age_flb_cat, 
                          fath_age_flb_cat)

# this counts each cross term as a separate test:
n_regs <- as.double(nrow(res_age_flb_mothers_cross))

res_age_flb_mothers_cross %>% 
      mutate(
        `Age at first live birth` = gsub("moth_age_birth_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, 
                                          "moth_age_flb_cat", dv = "n_sibs",
                                          data = famhist_bo1)
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs,
                      fill_direction = -1)

# this counts each cross term as a separate test:
n_regs <- as.double(nrow(res_age_flb_fathers_cross))

res_age_flb_fathers_cross %>% 
      mutate(
        `Age at first live birth` = gsub("fath_age_birth_cat(.*):.*", "\\1", term),
        `Age at first live birth` = add_n(`Age at first live birth`, 
                                          "fath_age_flb_cat", dv = "n_sibs",
                                          data = famhist_bo1)
      ) %>% 
      standard_ggplot(score_col = score_name, 
                      fill_col = `Age at first live birth`, n_regs = n_regs,
                      fill_direction = -1)

rm(famhist_bo1)

Figure \@ref(fig:plot-age-birth-parents) shows the regressions controlling for
either parent's age at their birth. Effect sizes are very similar, whether
controlling for father's or mother's age. As in the respondents' generation,
effect sizes are negatively correlated with the effect sizes from bivariate
regressions without the age at birth control (father's age at birth: $\rho$ 
`r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).
